---
title: "week2: Distances and the Border Zone"
format: html
editor: visual
author: 
  - name: Helen Flynn
  email: helen.flynn@colostate.edu
---

```{r}
# spatial data science
library(tidyverse)
library(sf)
library(units)

# Data
library(USAboundaries)
#had to run this to install USAboundaries: devtools::install_github("ropensci/USAboundariesData")
library(rnaturalearth)

# Visualization
library(gghighlight)
library(ggrepel)
library(knitr)
```

#Question 1

##1.1 Define a projection

```{r}
eqdc <- '+proj=eqdc +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs'
```

##1.2 Get USA state boundaries

```{r}
statedat <- USAboundaries::us_states(resolution = "low") %>% 
  filter(jurisdiction_type == "state",
         state_name != "Alaska",
          state_name != "Hawaii") %>% 
  st_transform(crs = eqdc)
```

##1.3 Get country boundaries for Mexico, the United States of America, and Canada
```{r}
countrydat <- countries110

nacountrydat <- countrydat %>% 
  filter(ADMIN %in% c("Mexico","United States of America","Canada")) %>% 
  st_as_sf() %>% 
  st_transform(crs = eqdc)
```

##1.4 Get city locations from the CSV file
```{r}
citydat <- read.csv(file = "C:/Users/hflynn19/OneDrive - Colostate/GEOGRAD/ESS523C/csu-ess523c/data/uscities.csv") %>% 
  filter(!state_id %in% c("AK","HI")) %>% 
  st_as_sf(coords = c("lng","lat"), crs = 4269) %>% 
  st_transform(crs = eqdc)
```

#Question 2
##2.1 Distance to USA border

```{r}
us_bound <- statedat %>% 
  st_union() %>% 
  st_cast("MULTILINESTRING")

library(flextable)
uscities_x <- citydat %>%
  st_filter(statedat) %>% 
  mutate(xusbound = st_distance(., us_bound))

maxdist_cities <- uscities_x %>%
  arrange(desc(xusbound)) %>%
  head(5)


maxdist_cities_table <- maxdist_cities %>%
  select(city, state_name, xusbound) %>%
  st_drop_geometry() %>% 
  flextable()
print(maxdist_cities_table)
  
```

##2.2 Distance to States
```{r}
us_bound_combine <- statedat %>% 
  st_combine() %>% 
  st_cast("MULTILINESTRING")

uscities_x <- uscities_x %>% 
   st_filter(statedat) %>% 
  mutate(xstatebound = st_distance(., us_bound_combine))

maxdist_states <- uscities_x %>%
  arrange(desc(xstatebound)) %>%
  head(5)
maxdist_states_table <- maxdist_states %>%
  select(city, state_name, xstatebound) %>%
  flextable()
print(maxdist_states_table)
```

##2.3 Distance to Mexico
```{r}
mex <- nacountrydat %>% 
  filter(GEOUNIT == "Mexico")

uscities_x <- uscities_x %>%
  mutate(xmexbound = st_distance(., mex))

maxdist_mex <- uscities_x %>%
  arrange(desc(xmexbound)) %>%
  head(5)
maxdist_mex_table <- maxdist_mex %>%
  select(city, state_name, xmexbound) %>%
  flextable()
print(maxdist_mex_table)
```

##2.4 Distance to Canada
```{r}
can <- nacountrydat %>% 
  filter(GEOUNIT == "Canada")

uscities_x <- uscities_x %>%
  mutate(xcanbound = st_distance(., can))

maxdist_can <- uscities_x %>%
  arrange(desc(xcanbound)) %>%
  head(5)
maxdist_can_table <- maxdist_can %>%
  select(city, state_name, xcanbound) %>%
  flextable()
print(maxdist_can_table)
```

#Question 3
##3.1 Data
```{r}
library(ggplot2)


ggplot() +
  geom_sf(data = uscities_x, aes(fill = population)) +
  gghighlight(max(population), max_highlight = 10,
              label_key = city)+
  # ggrepel::geom_label_repel(
  #   data = uscities_x,
  #   aes(label = city,
  #   stat = "geometry"),
  #   size = 3)+
  geom_sf(data = nacountrydat, fill = "NA")+
  geom_sf(data = us_bound, fill = "NA")+
  geom_sf(data = statedat, fill = "NA")

#not sure why this is not displaying correctly. Labels were causing errors as well. 

```
##3.2 City distance from the border
```{r}

```

##3.3 City distance from nearest state
##3.4 Equidistant boundary from Mexico and Canada

#Question 4
##4.1 Quantifing border zone
##4.2 Mapping border zone
##4.3 Instead of labeling the 10 most populous cites, label the most populous city in each state within the Danger Zone












